\MakeUppercase{The Structure of $k$-Planar Graphs}

### 2019-07-08 k-planar.tex
David's changes

\documentclass{patmorin}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amsthm}
\usepackage{graphicx}
\usepackage{enumerate}
\usepackage{amsfonts}
\usepackage{amsthm,mathtools}
\usepackage{pat}
\usepackage{paralist}
\usepackage{stmaryrd}

@@

@@

Here $\boxtimes$ is the strong product, and treewidth is an invariant that measures how `tree-like' a given graph is; see \cref{Tools} for formal definitions and see \cref{ProductExample} for an example.  So, loosely speaking, \cref{PlanarBasic} says that every planar graph is contained in the product of a tree-like graph and a path. This enables combinatorial results for graphs of bounded treewidth to be generalised for planar graphs (with different constants). \cref{kPlanarBasic} has been the key tool in solving several well-known open problems. In particular, \citet{dujmovic.joret.ea:planar} proved that planar graphs have bounded queue-number (resolving a conjecture of \citet{HLR92}), and \citet{dujmovic.esperet.ea:planar}  proved that planar graphs have bounded non-repetitive chromatic number (resolving a conjecture of \citet{AGHR-RSA02}).

\begin{figure}[!h]
\centering
\includegraphics{ProductExample}
\caption{Example of a strong product.}
\label{ProductExample}
\end{figure}

@@

\begin{thm}
\label{kPlanarBasic}
Every $k$-planar graph is a subgraph of $H\boxtimes P$, for some graph $H$ of treewidth at most $(4+o(1))k^5$ and for some path $P$. 
\end{thm}

This theorem and several related results are proved in \cref{Structure}. In particular, in \cref{sec-1-planar} we present a refinement of \cref{kPlanarBasic} in the case $k=1$. In \cref{sec-g-k-planar} we generalise \cref{kPlanarBasic} for graphs that have a drawing on an arbitrary surface with a bounded number of crossings per edge.

\cref{kPlanarBasic} (and its generalisations) have applications in diverse areas, including queue layouts  \citep{dujmovic.joret.ea:planar}, non-repetitive colouring  \citep{dujmovic.esperet.ea:planar}, and $p$-centred colouring  \citep{micek:personal}.
%, and graph encoding \citep{BGP}. 
These applications in the context of $k$-planar graphs are explored in \cref{Applications}. For example, we prove that $k$-planar graphs have bounded non-repetitive chromatic number (for fixed $k$). Prior to the recent work of 
\citet{dujmovic.esperet.ea:planar}, it was even open whether planar graphs have bounded non-repetitive chromatic number.

\cref{Characterisation} presents a rough characterisation of $k$-planar graphs in terms of an underlying planar graph. This is useful for showing that various families of graphs are $k$-planar. For example, we conclude that bounded powers of bounded-degree planar graphs are $k$-planar, for a suitable constant $k$. Thus 
\cref{kPlanarBasic} and the above applications are applicable to this class. These results also generalise for arbitrary surfaces.

\cref{Examples} presents several examples of $k$-planar graphs, including $k$-map graphs, $k$-string graphs, and $k$-nearest neighbour graphs. All of the above results are applicable to these examples.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Tools}
\label{Tools}

@@

While strong products enable an easy statement of \cref{PlanarBasic,kPlanarBasic}, to prove such theorems it is helpful to work with layerings and partitions, which we now introduce. A \emph{layering} of a graph $G$ is a sequence $\mathcal{L}=\langle V_0,V_1,\ldots\rangle$ such that $\{V_0,V_1,\ldots\}$ is a partition of $V(G)$ and for every edge $vw\in E(G)$, if $v\in V_i$ and $w\in V_j$ then $|j-i|\leq 1$.  For any partition $\mathcal{P}=\{S_1,\ldots,S_p\}$ of $V(G)$, a \emph{quotient graph} $H=G/\mathcal{P}$ has a $p$-element vertex set $V(H)=\{x_1,\ldots,x_p\}$ and $x_ix_j\in E(H)$ if and only if there exists an edge $vw\in E(G)$ such that $v\in S_i$ and $w\in S_j$. To highlight the importance of the quotient graph $H$, we call $\mathcal{P}$ an \emph{$H$-partition} and write this concisely as $\mathcal{P}=\{S_x : x\in V(H)\}$ so that each element of $\mathcal{P}$ is indexed by the vertex it creates in $H$.\footnote{An alternative definition of a layering of $G$ is an $H$-partition of $G$ where $H$ is a path.}  A layered $H$-partition $(\mathcal{L},\mathcal{P})$ of a graph $G$ consists of a layering $\mathcal{L}$ of $G$ and an $H$-partition $\mathcal{P}$ of $G$. The \emph{layered width} of $(\mathcal{L},\mathcal{P})$ is $\max\{|L\cap P|: L\in\mathcal{L},\, P\in\mathcal{P}\}$. These definitions relate to strong products as follows.

@@

\citet{dujmovic.joret.ea:planar} proved the following result, which with \cref{PartitionProduct,MakeWidth1}, implies \cref{PlanarBasic}.

@@

\begin{thm}[\citep{dujmovic.joret.ea:planar}]
\label{SurfacePartition}
Every graph of Euler genus $g$ is a subgraph of:
\begin{compactenum}[(a)]
\item  $H \boxtimes P \boxtimes K_{\max\{2g,1\}}$ for some graph $H$ of treewidth at most $9$ and for some path $P$;
\item  $H \boxtimes P \boxtimes K_{\max\{2g,3\}}$ for some graph $H$ of treewidth at most $4$ and for some path $P$;
\item $(K_{2g} + H )  \boxtimes P$ for some graph $H$ of treewidth at most $8$  and some path $P$. 
\end{compactenum}
Equivalently, every graph of Euler genus $g$ has:
\begin{compactenum}[(a)]
\item an $H$-partition with layered width at most $\max\{2g,1\}$ such that $H$ has treewidth at most $9$;
\item an $H$-partition with layered width at most $\max\{2g,3\}$ such that $H$ has treewidth at most $4$; 
\item an $H$-partition with layered width at most $1$ such that $H$ has treewidth at most $2g+7$. 
\end{compactenum}
%\textcolor{red}{eliminate some of these results?}
\end{thm}

\citet{dujmovic.joret.ea:planar} generalised \cref{SurfacePartition} further to show that for every apex\footnote{A graph $M$ is a \textit{minor} of a graph $G$ if a graph isomorphic to $M$ can be obtained from a subgraph of $G$ by contracting edges. A class $\mathcal{G}$ of graphs is \emph{minor-closed} if for every graph $G\in\mathcal{G}$, every minor of $G$ is in $\mathcal{G}$. A minor-closed class is \emph{proper} if it is not the class of all graphs. For example, for fixed $g\geq 0$, the class of graphs with Euler genus at most $g$ is a proper minor-closed class. A graph $G$ is $t$-apex if it contains a set $A$ of at most $t$ vertices such that $G-A$ is planar. A 1-apex graph is \emph{apex}.  A minor-closed class $\mathcal{G}$ is apex-minor-free if some apex graph is not in $\mathcal{G}$.}
 graph $X$, there is a constant $c$ such that every $X$-minor-free graph is a subgraph of $H\boxtimes P$ for some graph $H$ of treewidth at most $c$. Apex-minor-free graphs are the largest minor-closed class for which such a theorem holds \citep{dujmovic.joret.ea:planar}. However, the family of $k$-planar graphs is not minor-closed.  A graph $G'$ obtained from a $k$-planar graph $G$ by edge deletions and edge contractions may or may not be $k$-planar. Indeed, \citet{dujmovic.eppstein.ea:structure} construct 1-planar graphs that contain arbitrarily large complete graph minors. Our results for $k$-planar graphs are the first of this type for a non-minor-closed class.

The following is the main result of this paper; it shows that $k$-planar graphs admit $H$-partitions with bounded layered width, in which $H$ has bounded treewidth. With \cref{PartitionProduct,MakeWidth1}, this proves \cref{kPlanarBasic}. % with $c_k\leq(4+o(1))k^5$.

\begin{thm}
\label{k-planar}
Every $k$-planar graph has a layered $H$-partition of layered width at most $18k^2 + 30k$ in which $H$ has treewidth at most $\treewidth$.
\end{thm}

In the special case $k=1$ we obtain better constants and an additional property (planarity) of $H$:

@@

Prior to this work, the strongest structural description of $k$-planar graphs was in terms of layered treewidth, which we now define.  A \emph{layered tree-decomposition} $(\mathcal{L},\mathcal{K})$ consists of a layering $\mathcal{L}$ and a tree-decomposition $\mathcal{K}$ of $G$. The layered width of $(\mathcal{L},\mathcal{K})$ is $\max\{|L\cap B|: L\in \mathcal{L},\, B\in \mathcal{K}\}$.  The \emph{layered treewidth} of $G$ is the minimum layered width of any layered tree-decomposition of $G$. \citet{dujmovic.morin.ea:layered} proved that planar graphs have layered treewidth at most 3, that graphs of Euler genus $g$ have layered treewidth at most $2g+3$, and more generally that a minor-closed class has bounded layered treewidth if and only if it excludes some apex graph. \citet{dujmovic.eppstein.ea:structure} show that $k$-planar graphs have layered treewidth at most $6(k+1)$.

If a graph has a partition with bounded layered width, where the quotient has bounded treewidth, then the graph has bounded layered treewidth. In particular, suppose that an $n$-vertex graph $G$ has an $H$-partition with layered width at most $\ell$ such that $H$ has treewidth at most $k$. \citet{dujmovic.joret.ea:planar} proved that $G$ has layered treewidth at most $(k+1)\ell$. So any property that holds for graphs of bounded layered treewidth also holds for $G$. For example, Norin proved that every $n$-vertex graph with layered treewidth at most $t$ has treewidth less than $2\sqrt{tn}$ (see \citep{dujmovic.morin.ea:layered}). Thus, $G$ has treewidth at most $2\sqrt{(k+1)\ell n}$. This in turn leads to a $O(\sqrt{(k+1)\ell n})$ balanced separator theorem for such graphs \citep{RS-II}. So, in the case of $k$-planar graphs, \cref{k-planar} implies an analogue of the separator theorem of \citet{FP08} mentioned in the introduction, albeit with worse dependence on $k$. What sets layered partitions apart from layered treewidth is that they lead to constant upper bounds on the queue-number  and non-repetitive chromatic number, whereas for both these parameters, the best known upper bound obtainable via layered treewidth is $O(\log n)$; see \cref{Applications}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Structure}
\label{Structure}

The purpose of this section is to prove \cref{k-planar} and related structural results.  We begin with a more general abstraction called $(k,d)$-shortcut systems and prove a structural result, \cref{modern-major-general}, for graphs derived from $(k,d)$-shortcut systems.  The results on $k$-planar (\cref{k-planar}) and $(g,k)$-planar graphs (\cref{g-k-planar}) can then be obtained from \cref{modern-major-general} and existing results for planar and bounded-genus graphs.

\subsection{Shortcut Systems}

For a rooted tree $T$ and any node $x\in V(T)$, we define $T_x := T[\{y\in V(T):\mbox{$x$ is a $T$-ancestor of $y$}\}]$ to be the maximal subtree of $T$ rooted at $x$.  We begin with a fairly standard lemma about normalized tree decompositions:

@@

\subsubsection{Generalized Tripod Partitions}

@@

\subsubsection{Shortcut Systems}

@@

A collection $\mathcal{P}$ of paths in a graph $G$ is a \emph{$(k,d)$-shortcut system} (for $G$) if every path in $\mathcal{P}$ has length at most $k$ and if, for every $v\in V(G)$, the number of paths in $\mathcal{P}$ that use $v$ as an internal vertex is at most $d$. Each path in $P\in\mathcal{P}$ is called a \emph{shortcut}; if $P$ has endpoints $v$ and $w$ then it is a \emph{$vw$-shortcut}. Given a graph $G$ and a $(k,d)$-shortcut system $\mathcal{P}$ for $G$, we use $G^{\mathcal{P}}$ to denote the supergraph of $G$ obtained by adding the edge $vw$ for each $vw$-shortcut in $\mathcal{P}$.

In this section we will show how, given a $(k,d)$-shortcut system $\mathcal{P}$, a layered $H$-partition of $G$ can be used to obtain a layered $I$-partition of $G^{\mathcal{P}}$ where the layered width of the partition does not increase dramatically and the treewidth of $I$ is not much more than the treewidth of $H$.

@@

% Together, \clmref{general-width} and \clmref{general-bag-size} imply the following generalization of \thmref{k-planar}:

\begin{thm}\label{modern-major-general}
  Let $G$ be a graph having a layered $H$-partition of layered width $\ell$ in which $H$ has treewidth at most $t$ and let $\mathcal{P}$ be a $(k,d)$-shortcut system for $G$.  Then $G^{\mathcal{P}}$ has a layered $I$-partition of layered width at most $d\ell(k^3+k^2+4k)$ where the treewidth of $H$ is at most $\binom{k+t}{t}-1$.
\end{thm}

\begin{proof}
  The proof is almost immediate, except that the layering $\mathcal{L}$ of $G$ may not be a valid layering of $G^{\mathcal{P}}$.  In particular, $G^{\mathcal{P}}$ may contain an edge $vw$ with $v\in V_i$ and $w\in V_{i+j}$ for any $j\in\{0,\ldots,k\}$.  To resolve this, we use a new layering $\mathcal{L}':=\langle V_0',\ldots,V_h'\rangle$ in which $V_i'=\bigcup_{j=ki}^{ki+k-1} V_i$.  This increases the thickness given by \clmref{general-width} from $d\ell(k^2+k+4)$ to $d\ell(k^3+k^2+4k)$.
\end{proof}

\subsection{$k$-Planar Graphs}
\label{sec-k-planar}

@@

Let $G_0$ be the plane graph obtained by adding a dummy vertex at each crossing in $G$. In this way, each edge $vw\in E(G)$ corresponds naturally to a path $P_{vw}$ of length at most $k+1$ in $G_0$.  Take $\mathcal{P}=\{P_{vw}: vw\in E(G)\}$ and observe that $\mathcal{P}$ is a $(k+1,2)$-shortcut for $G_0$ and that $G_0^{\mathcal{P}}\supseteq G$.  Specifically, $G_0^{\mathcal{P}}$ contains every edge and vertex of $G$ as well as the dummy vertices in $V(G_0)\setminus V(G)$ and their incident edges.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{$(g,k)$-Planar Graphs}
\label{sec-g-k-planar}

The definition of $k$-planar graphs naturally generalises for other surfaces. A graph $G$ drawn on a surface $\Sigma$ is $(\Sigma,k)$-plane if every edge of $G$ is involved in at most $k$ crossings.  A graph $G$ is $(g,k)$-planar if it is isomorphic to some $(\Sigma,k)$-plane graph, for some surface $\Sigma$ with Euler genus at most $g$.

>>>>

\documentclass{patmorin}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{amsmath,wrapfig}
\usepackage{amsfonts}
\usepackage{amsthm}
\usepackage{graphicx}
\usepackage{enumerate}
\usepackage{amsfonts}
\usepackage{amsthm,mathtools}
\usepackage{pat}
\usepackage{paralist}
\usepackage{stmaryrd}

@@

\newcommand{\PP}{\mathcal{P}}

@@

Here $\boxtimes$ is the strong product, and treewidth is an invariant that measures how `tree-like' a given graph is; see \cref{Tools} for formal definitions and see \cref{ProductExample} for an example.  
So, loosely speaking, \cref{PlanarBasic} says that every planar graph is contained in the product of a tree-like graph and a path. This enables combinatorial results for graphs of bounded 
\begin{wrapfigure}{r}{0.5\textwidth} 
\centering
\includegraphics{ProductExample}
\caption{Example of a strong product.}
\label{ProductExample}
\end{wrapfigure}
treewidth to be generalised for planar graphs (with different constants). \cref{kPlanarBasic} has been the key tool in solving several well-known open problems. In particular, \citet{dujmovic.joret.ea:planar} proved that planar graphs have bounded queue-number (resolving a conjecture of \citet{HLR92}), and \citet{dujmovic.esperet.ea:planar}  proved that planar graphs have bounded non-repetitive chromatic number (resolving a conjecture of \citet{AGHR-RSA02}).

@@

\begin{thm}
\label{kPlanarBasic}
Every $k$-planar graph is a subgraph of $H\boxtimes P$, for some graph $H$ of treewidth $O(k^5)$ and for some path $P$. 
\end{thm}

We actually prove a substantially stronger result than \cref{kPlanarBasic} using the following definition. A collection $\mathcal{P}$ of paths in a graph $G$ is a \emph{$(k,d)$-shortcut system} (for $G$) if:
\begin{compactitem}
\item every path in $\mathcal{P}$ has length at most $k$, and
\item for every $v\in V(G)$, the number of paths in $\mathcal{P}$ that use $v$ as an internal vertex is at most $d$.
\end{compactitem} 
Each path $P\in\mathcal{P}$ is called a \emph{shortcut}; if $P$ has endpoints $v$ and $w$ then it is a \emph{$vw$-shortcut}. Given a graph $G$ and a $(k,d)$-shortcut system $\mathcal{P}$ for $G$, let $G^{\mathcal{P}}$ denote the supergraph of $G$ obtained by adding the edge $vw$ for each $vw$-shortcut in $\mathcal{P}$.

This definition is related to $k$-planarity because of the following observation

\begin{obs}
\label{AddDummy}
Every $k$-planar graph is isomorphic to $G^\PP$ for some planar graph $G$ and some $(k+1,2)$-shortcut system $\PP$ for $G$. 
\end{obs}

The proof of \cref{AddDummy} is trivial: Given a $k$-planar graph $G'$, create a planar graph $G$ by adding a dummy vertex at each crossing point. For each edge $vw\in E(G')\setminus E(G)$ there is a path $P$ in $G$ between $v$ and $w$ of length at most $k+1$. Let $\PP$ be the set of such paths $P$. For each vertex $v$ of $G$, at most two paths in $\PP$ use $v$ as an internal vertex (since no original vertex of $G'$ is an internal vertex of a path in $\PP$). Thus $\PP$ is a $(k+1,2)$-shortcut system for $G$, such that $G'=G^\PP$.

We prove the following theorem.

\begin{thm}
\label{ShortcutProduct}
Let $G$ be a subgraph of $H\boxtimes P$, for some graph $H$ of treewidth at most $t$ and for some path $P$. 
Let $\PP$ be a $(k,d)$-shortcut system for $G$. Then $G^\PP$ is a subgraph of $J\boxtimes P$ for some graph $J$ of treewidth at most $d(k^3+k^2+4k)\binom{k+t}{t}-1$ and some path $P$. 
\end{thm}

\cref{ShortcutProduct,AddDummy} immediately imply \cref{kPlanarBasic} with $O(k^{11})$ instead of $O(k^5)$. Some further observations reduce the bound to $O(k^5)$; see \cref{Tools}. All of these results  generalise for graphs that have a drawing on an arbitrary surface with a bounded number of crossings per edge.

\cref{ShortcutProduct} leads to several other results of interest. Here is one example. The \emph{$k$-th power} of a graph $G$ is the graph $G^k$ with vertex set $V(G^k):=V(G)$, where $vw\in E(G^k)$ if and only if $\dist_G(v,w)\leq k$. If $G$ has maximum degree $\Delta$, then $G^k = G^\PP$ for some $(k,2k\Delta^{k})$-shortcut system $\PP$; see \cref{PowerShortcut}. \cref{kPlanarBasic,ShortcutProduct} then imply:

\begin{thm}
\label{kPowerBasic}
For every planar graph $G$ with maximum degree $\Delta$ and for every integer $k\geq 1$, $G^k$ is a subgraph of $H\boxtimes P$, for some graph $H$ of treewidth at most $O( k^7 \Delta^{k})$ and for some path $P$. \note{DW}{Improve to $k^6\Delta^{k}$?}
\end{thm}

%This theorem and several related results are proved in \cref{Structure}. In particular, in \cref{sec-1-planar} we present a refinement of \cref{kPlanarBasic} in the case $k=1$. In \cref{sec-g-k-planar} we generalise \cref{kPlanarBasic} for

These theorems have applications in diverse areas, including queue layouts  \citep{dujmovic.joret.ea:planar}, non-repetitive colouring  \citep{dujmovic.esperet.ea:planar}, and $p$-centred colouring  \citep{micek:personal}, 
%, and graph encoding \citep{BGP}. 
which we explore in \cref{Applications}. For example, we prove that $k$-planar graphs have bounded non-repetitive chromatic number (for fixed $k$). Prior to the recent work of \citet{dujmovic.esperet.ea:planar}, it was even open whether planar graphs have bounded non-repetitive chromatic number.

%\cref{Characterisation} presents a rough characterisation of $k$-planar graphs in terms of an underlying planar graph. This is useful for showing that various families of graphs are $k$-planar. For example, we conclude that bounded powers of bounded-degree planar graphs are $k$-planar, for a suitable constant $k$. Thus  \cref{kPlanarBasic} and the above applications are applicable to this class. These results also generalise for arbitrary surfaces.

\cref{Examples} presents several examples of $k$-planar graphs, including graph powers, map graphs, string graphs, and nearest neighbour graphs. All of the above results are applicable to these examples.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Layerings, Decompositions and Partitions}
\label{Tools}

@@

While strong products enable an easy statement of \cref{PlanarBasic,kPlanarBasic}, to prove such theorems it is helpful to work with layerings and partitions, which we now introduce. A \emph{layering} of a graph $G$ is a sequence $\mathcal{L}=\langle V_0,V_1,\ldots\rangle$ such that $\{V_0,V_1,\ldots\}$ is a partition of $V(G)$ and for every edge $vw\in E(G)$, if $v\in V_i$ and $w\in V_j$ then $|j-i|\leq 1$.  For any partition $\mathcal{P}=\{S_1,\ldots,S_p\}$ of $V(G)$, a \emph{quotient graph} $H=G/\mathcal{P}$ has a $p$-element vertex set $V(H)=\{x_1,\ldots,x_p\}$ and $x_ix_j\in E(H)$ if and only if there exists an edge $vw\in E(G)$ such that $v\in S_i$ and $w\in S_j$. To highlight the importance of the quotient graph $H$, we call $\mathcal{P}$ an \emph{$H$-partition} and write this concisely as $\mathcal{P}=\{S_x : x\in V(H)\}$ so that each element of $\mathcal{P}$ is indexed by the vertex it creates in $H$.\footnote{An alternative definition of a layering of $G$ is an $H$-partition of $G$ where $H$ is a path.}  A layered $H$-partition $(\mathcal{L},\mathcal{P})$ of a graph $G$ consists of a layering $\mathcal{L}$ of $G$ and an $H$-partition $\mathcal{P}$ of $G$. The \emph{layered width} of $(\mathcal{L},\mathcal{P})$ is $\max\{|L\cap P|: L\in\mathcal{L},\, P\in\mathcal{P}\}$.

\citet{dujmovic.joret.ea:planar} introduced the study of partitions with bounded layered width such that the quotient has bounded treewidth. They defined a class $\mathcal{G}$ of graphs to \emph{admit bounded layered partitions} if there exist $t,\ell\in\mathbb{N}$ such that every graph $G\in \mathcal{G}$ has an $H$-partition with layered width at most $\ell$ such that $H$ has treewidth at most $t$.

These definitions relate to strong products as follows.

@@

\citet{dujmovic.joret.ea:planar} proved the following result, which with \cref{PartitionProduct}, implies \cref{PlanarBasic}.

@@

\begin{thm}[\citep{dujmovic.joret.ea:planar}]
\label{SurfacePartition}
Every graph of Euler genus $g$ is a subgraph of:
\begin{compactenum}[(a)]
\item  $H \boxtimes P \boxtimes K_{\max\{2g,1\}}$ for some graph $H$ of treewidth at most $9$ and for some path $P$;
\item  $H \boxtimes P \boxtimes K_{\max\{2g,3\}}$ for some graph $H$ of treewidth at most $4$ and for some path $P$;
\item $(K_{2g} + H )  \boxtimes P$ for some graph $H$ of treewidth at most $8$  and some path $P$. 
\end{compactenum}
Equivalently, every graph of Euler genus $g$ has:
\begin{compactenum}[(a)]
\item an $H$-partition with layered width at most $\max\{2g,1\}$ such that $H$ has treewidth at most $9$;
\item an $H$-partition with layered width at most $\max\{2g,3\}$ such that $H$ has treewidth at most $4$; 
\item an $H$-partition with layered width at most $1$ such that $H$ has treewidth at most $2g+7$. 
\end{compactenum}
\textcolor{red}{eliminate some of these results? just keep (b)? }
\end{thm}

\citet{dujmovic.joret.ea:planar} generalised \cref{SurfacePartition} further to show that for every apex\footnote{A graph $M$ is a \textit{minor} of a graph $G$ if a graph isomorphic to $M$ can be obtained from a subgraph of $G$ by contracting edges. A class $\mathcal{G}$ of graphs is \emph{minor-closed} if for every graph $G\in\mathcal{G}$, every minor of $G$ is in $\mathcal{G}$. A minor-closed class is \emph{proper} if it is not the class of all graphs. For example, for fixed $g\geq 0$, the class of graphs with Euler genus at most $g$ is a proper minor-closed class. A graph $G$ is $t$-apex if it contains a set $A$ of at most $t$ vertices such that $G-A$ is planar. A 1-apex graph is \emph{apex}.  A minor-closed class $\mathcal{G}$ is apex-minor-free if some apex graph is not in $\mathcal{G}$.}
 graph $X$, there is a constant $c$ such that every $X$-minor-free graph is a subgraph of $H\boxtimes P$ for some graph $H$ of treewidth at most $c$. Apex-minor-free graphs are the largest minor-closed class that admit bounded layered partitions~\citep{dujmovic.joret.ea:planar}. However, the family of $k$-planar graphs is not minor-closed.  A graph $G'$ obtained from a $k$-planar graph $G$ by edge deletions and edge contractions may or may not be $k$-planar. Indeed, \citet{dujmovic.eppstein.ea:structure} construct 1-planar graphs that contain arbitrarily large complete graph minors. Our results for $k$-planar graphs are the first of this type for a non-minor-closed class.

\section{Our Results}

The following result is the main theorem of the paper. Loosely speaking, it shows that if a graph $G$ admits bounded layered partitions, then so to does $G^\PP$ for every $(k,d)$-shortcut system $\PP$.

\begin{thm}\label{modern-major-general}
  Let $G$ be a graph having a layered $H$-partition of layered width $\ell$ in which $H$ has treewidth at most $t$ and let $\PP$ be a $(k,d)$-shortcut system for $G$.  Then $G^\PP$ has a layered $J$-partition of layered width at most $d\ell(k^3+k^2+4k)$ for some graph $J$ with treewidth at most $\binom{k+t}{t}-1$.
\end{thm}

\cref{modern-major-general,PartitionProduct,MakeWidth1} immediately imply \cref{ShortcutProduct} in the introduction.

The following result shows that $k$-planar graphs admit bounded layered partitions. With \cref{PartitionProduct,MakeWidth1}, this proves \cref{kPlanarBasic}. % with $c_k\leq(4+o(1))k^5$.

\begin{thm}
\label{k-planar}
Every $k$-planar graph has a layered $H$-partition of layered width at most $18k^2 + 30k$ in which $H$ has treewidth at most $\binom{k+4}{3}-1$. %$\treewidth$.
\end{thm}

\cref{PlanarPartition,modern-major-general} imply \cref{k-planar} with a weaker bound on the layered width. We prove the stronger bound in \cref{sec-k-planar}. Also note that \cref{k-planar,PartitionProduct} imply \cref{kPlanarBasic} with the $O(k^5)$ bound on the treewidth of $H$, as promised.

In the special case $k=1$ we obtain better constants and an additional property (planarity) of $H$ (see \cref{sec-1-planar} for the proof):

@@

The definition of $k$-planar graphs naturally generalises for other surfaces. A graph $G$ drawn on a surface $\Sigma$ is $(\Sigma,k)$-plane if every edge of $G$ is involved in at most $k$ crossings.  A graph $G$ is $(g,k)$-planar if it is isomorphic to some $(\Sigma,k)$-plane graph, for some surface $\Sigma$ with Euler genus at most $g$. 
 %\citet[Theorem~22]{dujmovic.joret.ea:planar} show that every genus-$g$ graph has a layered $H$-partition of layered width at most $\max\{2g,3\}$ such that $H$ has treewidth at most 4. 
\cref{AddDummy} immediately generalises as follows:

\begin{obs}
\label{gAddDummy}
Every $(g,k)$-planar graph is isomorphic to $G^\PP$ for some graph $G$ of Euler genus $g$ and some $(k+1,2)$-shortcut system $\PP$ for $G$. 
\end{obs}

We prove that $(g,k)$-planar graphs admits bounded layered partitions.

\begin{thm}
\label{g-k-planar}
Every $(g,k)$-planar graph has a layered $H$-partition of layered width at most 
$\max\{2g,3\}\cdot(6k^2 + 10k)$ in which $H$ has treewidth at most $\binom{k+5}{4}-1$.
\end{thm}

Note that \cref{SurfacePartition}(b) and \cref{modern-major-general} imply \cref{k-planar} with a weaker bound on the layered width. We prove the stronger bound in \cref{sec-k-planar}.

Finally, we state the following corollary of \cref{modern-major-general,MakeWidth1,PartitionProduct} and the above-mentioned theorem for apex-minor-free graphs.  This is the strongest result that follows from our work and that of \citet{dujmovic.joret.ea:planar}.

\begin{thm}
For every apex graph $X$ and for all integers $k,d\geq 1$, there is an integer $c$ such that for every $X$-minor-free graph $G$ and for every $(k,d)$-shortcut system $\PP$ for $G$,  $G^\PP$ has an $H$-partition of layered width 1 such that $H$ has treewidth at most $c$; that is $G^\PP \subseteq H \boxtimes P$ for some path $P$. 
\end{thm}

%%%%%%%%%%%%%%%%
\subsection{Prior Work}
\label{PriorWork}

\note{DW}{should this subsection be moved to the end of \cref{Tools}?}

Prior to this work, the strongest structural description of $k$-planar or $(g,k)$-planar graphs was in terms of layered treewidth, which we now define.  A \emph{layered tree-decomposition} $(\mathcal{L},\mathcal{K})$ consists of a layering $\mathcal{L}$ and a tree-decomposition $\mathcal{K}$ of $G$. The layered width of $(\mathcal{L},\mathcal{K})$ is $\max\{|L\cap B|: L\in \mathcal{L},\, B\in \mathcal{K}\}$.  The \emph{layered treewidth} of $G$ is the minimum layered width of any layered tree-decomposition of $G$. \citet{dujmovic.morin.ea:layered} proved that planar graphs have layered treewidth at most 3, that graphs of Euler genus $g$ have layered treewidth at most $2g+3$, and more generally that a minor-closed class has bounded layered treewidth if and only if it excludes some apex graph. \citet{dujmovic.eppstein.ea:structure} show that every $k$-planar graph has layered treewidth at most $6(k+1)$, and more generally that 
every $(g,k)$-planar graph has layered treewidth at most $(4g+6)(k+1)$.

If a graph class admits bounded layered partitions, then it also has bounded layered treewidth. In particular, suppose that an $n$-vertex graph $G$ has an $H$-partition with layered width at most $\ell$ such that $H$ has treewidth at most $k$. \citet{dujmovic.joret.ea:planar} proved that $G$ has layered treewidth at most $(k+1)\ell$. So any property that holds for graphs of bounded layered treewidth also holds for $G$. For example, Norin proved that every $n$-vertex graph with layered treewidth at most $t$ has treewidth less than $2\sqrt{tn}$ (see \citep{dujmovic.morin.ea:layered}). Thus, $G$ has treewidth at most $2\sqrt{(k+1)\ell n}$. This in turn leads to a $O(\sqrt{(k+1)\ell n})$ balanced separator theorem for such graphs \citep{RS-II}. So, in the case of $k$-planar graphs, \cref{k-planar} implies an analogue of the separator theorem of \citet{FP08} mentioned in the introduction, albeit with worse dependence on $k$. What sets layered partitions apart from layered treewidth is that they lead to constant upper bounds on the queue-number  and non-repetitive chromatic number, whereas for both these parameters, the best known upper bound obtainable via layered treewidth is $O(\log n)$; see \cref{Applications}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Shortcut Systems}
\label{Structure}

The purpose of this section is to prove our main result, \cref{modern-major-general}. This theorem shows how, given a $(k,d)$-shortcut system $\mathcal{P}$ of a graph $G$, a layered $H$-partition of $G$ can be used to obtain a layered $J$-partition of $G^{\mathcal{P}}$ where the layered width  does not increase dramatically and the treewidth of $J$ is not much more than the treewidth of $H$.  Our main results for $k$-planar graphs (\cref{k-planar}) and $(g,k)$-planar graphs (\cref{g-k-planar}) follow; see \cref{sec-k-planar}.

We begin with a fairly standard lemma about normalized tree decompositions. 
For a rooted tree $T$ and any node $x\in V(T)$, let  
$$T_x := T[\{y\in V(T):\mbox{$x$ is a $T$-ancestor of $y$}\}]$$ 
be the maximal subtree of $T$ rooted at $x$.

@@

%\subsection{Generalized Tripod Partitions}

@@

@@

%A collection $\mathcal{P}$ of paths in a graph $G$ is a \emph{$(k,d)$-shortcut system} (for $G$) if every path in $\mathcal{P}$ has length at most $k$ and if, for every $v\in V(G)$, the number of paths in $\mathcal{P}$ that use $v$ as an internal vertex is at most $d$. Each path in $P\in\mathcal{P}$ is called a \emph{shortcut}; if $P$ has endpoints $v$ and $w$ then it is a \emph{$vw$-shortcut}. Given a graph $G$ and a $(k,d)$-shortcut system $\mathcal{P}$ for $G$, we use $G^{\mathcal{P}}$ to denote the supergraph of $G$ obtained by adding the edge $vw$ for each $vw$-shortcut in $\mathcal{P}$.

@@

% Together, \clmref{general-width} and \clmref{general-bag-size} imply the following generalization of \thmref{k-planar}:
We are now ready to prove the main result of the paper.

\begin{proof}[Proof of \cref{modern-major-general}] 
  The proof is almost immediate, except that the layering $\mathcal{L}$ of $G$ may not be a valid layering of $G^{\mathcal{P}}$.  In particular, $G^{\mathcal{P}}$ may contain an edge $vw$ with $v\in V_i$ and $w\in V_{i+j}$ for any $j\in\{0,\ldots,k\}$.  To resolve this, we use a new layering $\mathcal{L}':=\langle V_0',\ldots,V_h'\rangle$ in which $V_i'=\bigcup_{j=ki}^{ki+k-1} V_i$.  This increases the layered width given by \clmref{general-width} from $d\ell(k^2+k+4)$ to $d\ell(k^3+k^2+4k)$.
\end{proof}

%%%%%%%%%%%%%%%%%%%
\section{\boldmath $k$-Planar Graphs}
\label{sec-k-planar}

@@

Let $G_0$ be the plane graph obtained by adding a dummy vertex at each crossing in $G$. In this way, each edge $vw\in E(G)$ corresponds naturally to a path $P_{vw}$ of length at most $k+1$ in $G_0$.  Let $\PP := \{P_{vw}: vw\in E(G)\}$. Observe that $\PP$ is a $(k+1,2)$-shortcut for $G_0$ and that $G_0^{\mathcal{P}}\supseteq G$.  Specifically, $G_0^{\mathcal{P}}$ contains every edge and vertex of $G$ as well as the dummy vertices in $V(G_0)\setminus V(G)$ and their incident edges.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{\boldmath  $(g,k)$-Planar Graphs}
\label{sec-g-k-planar}
